---
title: "Performance Benchmarking: Block-Based Processing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance Benchmarking: Block-Based Processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette demonstrates a **block-based processing** pattern for computing
bioclimatic variables over large climate datasets.  Rather than operating on
one pixel at a time (slow) or loading everything into memory at once
(impractical for global data), you can split data into blocks of rows,
process each block with vectorised R, and reassemble the results.

We benchmark four approaches:

1. **Scalar loop** – call `bioclim()` once per pixel in a `for` loop.
2. **`apply()` over rows** – call `bioclim()` via `apply()`.
3. **Block processing** – split into chunks, process each chunk with
   `apply()`, and bind results.
4. **`vapply()` loop** – call `bioclim()` via `vapply()`.

```{r setup}
library(rxbioclim)
```

## Generating synthetic climate data

We create a matrix of synthetic monthly climate values for `N` pixels.  Each
row is one pixel; columns are months.

```{r synthetic-data}
generate_climate <- function(n_pixels, seed = 42) {
  set.seed(seed)

  # Base annual mean temperature, uniformly distributed across latitudes
  tas_base <- runif(n_pixels, -5, 25)

  # Seasonal amplitude (larger at higher latitudes)
  amplitude <- runif(n_pixels, 2, 15)

  month_phase <- seq(0, 2 * pi, length.out = 13)[-13]

  TAS <- t(outer(amplitude, sin(month_phase), "*")) +
         matrix(tas_base, nrow = 12, ncol = n_pixels, byrow = TRUE)
  TAS <- t(TAS)  # n_pixels × 12

  TASMAX <- TAS + matrix(runif(n_pixels * 12, 2, 8), nrow = n_pixels)
  TASMIN <- TAS - matrix(runif(n_pixels * 12, 2, 8), nrow = n_pixels)

  # Precipitation: peaks in one season, roughly anti-correlated with temp
  pr_base <- runif(n_pixels, 10, 200)
  PR <- abs(
    t(outer(pr_base, sin(month_phase + pi / 2), "*")) +
      matrix(pr_base, nrow = 12, ncol = n_pixels, byrow = TRUE)
  )
  PR <- t(PR)  # n_pixels × 12
  PR[PR < 0] <- 0

  list(TAS = TAS, TASMAX = TASMAX, TASMIN = TASMIN, PR = PR)
}

N <- 1000  # number of pixels for benchmarking
clim <- generate_climate(N)
```

## Approach 1: Scalar `for` loop

```{r scalar-loop}
bioclim_loop <- function(TAS, TASMAX, TASMIN, PR) {
  n <- nrow(TAS)
  out <- matrix(NA_real_, nrow = n, ncol = 19)
  for (i in seq_len(n)) {
    out[i, ] <- bioclim(TAS[i, ], TASMAX[i, ], TASMIN[i, ], PR[i, ])
  }
  colnames(out) <- paste0("bio", sprintf("%02d", 1:19))
  out
}
```

## Approach 2: `apply()` over rows

```{r apply-rows}
bioclim_apply <- function(TAS, TASMAX, TASMIN, PR) {
  t(vapply(seq_len(nrow(TAS)), function(i) {
    bioclim(TAS[i, ], TASMAX[i, ], TASMIN[i, ], PR[i, ])
  }, numeric(19)))
}
```

## Approach 3: Block processing

Block processing mimics how raster packages (terra, stars) handle large
datasets: data is read, processed, and written in fixed-size blocks to limit
peak memory usage.

```{r block-processing}
bioclim_blocks <- function(TAS, TASMAX, TASMIN, PR, block_size = 100) {
  n <- nrow(TAS)
  n_blocks <- ceiling(n / block_size)
  results <- vector("list", n_blocks)

  for (b in seq_len(n_blocks)) {
    row_start <- (b - 1L) * block_size + 1L
    row_end   <- min(b * block_size, n)
    idx <- row_start:row_end

    block_result <- t(vapply(idx, function(i) {
      bioclim(TAS[i, ], TASMAX[i, ], TASMIN[i, ], PR[i, ])
    }, numeric(19)))
    results[[b]] <- block_result
  }

  out <- do.call(rbind, results)
  colnames(out) <- paste0("bio", sprintf("%02d", 1:19))
  out
}
```

## Approach 4: `vapply()` loop

```{r vapply-loop}
bioclim_vapply <- function(TAS, TASMAX, TASMIN, PR) {
  t(vapply(seq_len(nrow(TAS)), function(i) {
    bioclim(TAS[i, ], TASMAX[i, ], TASMIN[i, ], PR[i, ])
  }, numeric(19)))
}
```

## Correctness check

All four approaches should produce identical results (up to floating-point
precision):

```{r correctness}
r1 <- bioclim_loop(clim$TAS, clim$TASMAX, clim$TASMIN, clim$PR)
r2 <- bioclim_apply(clim$TAS, clim$TASMAX, clim$TASMIN, clim$PR)
r3 <- bioclim_blocks(clim$TAS, clim$TASMAX, clim$TASMIN, clim$PR)
r4 <- bioclim_vapply(clim$TAS, clim$TASMAX, clim$TASMIN, clim$PR)

stopifnot(
  all.equal(r1, r2, check.attributes = FALSE),
  all.equal(r1, r3, check.attributes = FALSE),
  all.equal(r1, r4, check.attributes = FALSE)
)
cat("All approaches produce identical results.\n")
```

## Benchmark

```{r benchmark}
times <- list(
  loop   = system.time(bioclim_loop(clim$TAS, clim$TASMAX, clim$TASMIN, clim$PR)),
  apply  = system.time(bioclim_apply(clim$TAS, clim$TASMAX, clim$TASMIN, clim$PR)),
  blocks = system.time(bioclim_blocks(clim$TAS, clim$TASMAX, clim$TASMIN, clim$PR)),
  vapply = system.time(bioclim_vapply(clim$TAS, clim$TASMAX, clim$TASMIN, clim$PR))
)

timing_df <- data.frame(
  approach   = names(times),
  elapsed_s  = sapply(times, `[[`, "elapsed"),
  row.names  = NULL
)
timing_df$pixels_per_sec <- round(N / timing_df$elapsed_s)
print(timing_df)
```

## Block size sensitivity

The optimal block size depends on your system's cache size and memory
bandwidth.  The plot below shows elapsed time as a function of block size
(run with N = 500 pixels for speed):

```{r block-size-sensitivity}
N_small <- 500
clim_s  <- generate_climate(N_small, seed = 7)

block_sizes  <- c(10, 25, 50, 100, 250, 500)
block_times  <- sapply(block_sizes, function(bs) {
  system.time(
    bioclim_blocks(clim_s$TAS, clim_s$TASMAX, clim_s$TASMIN, clim_s$PR,
                   block_size = bs)
  )[["elapsed"]]
})

sens_df <- data.frame(block_size = block_sizes, elapsed_s = block_times)
print(sens_df)
```

Typical observation: very small blocks (< 25) incur per-block overhead,
while very large blocks do not meaningfully improve throughput because the
bottleneck shifts to per-pixel computation inside `bioclim()`.  A block
size of 50–200 is usually a reasonable default.

## Scaling to large datasets

For truly large rasters (e.g., global 0.5° grids ≈ 260 000 land pixels),
consider:

1. **Pre-allocating output** – use `matrix(NA_real_, nrow = N, ncol = 19)`
   before the loop and fill rows in-place.
2. **Parallel processing** – wrap the block loop with `parallel::mclapply()`
   or `foreach` + `doParallel` to use multiple cores.
3. **Streaming I/O** – read blocks from disk with `terra::readValues()` or
   `stars::read_stars()`, process with rxbioclim, and write results back
   with `terra::writeValues()`.

A sketch of a parallel block loop:

```r
library(parallel)

bioclim_parallel_blocks <- function(TAS, TASMAX, TASMIN, PR,
                                    block_size = 200, n_cores = 2L) {
  n <- nrow(TAS)
  n_blocks <- ceiling(n / block_size)

  block_ids <- split(seq_len(n), ceiling(seq_len(n) / block_size))

  results <- mclapply(block_ids, function(idx) {
    t(vapply(idx, function(i) {
      bioclim(TAS[i, ], TASMAX[i, ], TASMIN[i, ], PR[i, ])
    }, numeric(19)))
  }, mc.cores = n_cores)

  out <- do.call(rbind, results)
  colnames(out) <- paste0("bio", sprintf("%02d", 1:19))
  out
}
```
