---
title: "Getting Started with xbioclim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with xbioclim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**xbioclim** computes the 19 standard bioclimatic variables (BIO01–BIO19)
from monthly climate data, following the
[WorldClim](https://www.worldclim.org/data/bioclim.html) specification.  It is
an R implementation of the
[xbioclim](https://github.com/alrobles/xbioclim) C++ library and exposes
both a *unified* function (`bioclim()`) and *individual* variable functions
(`bio01()` … `bio19()`).

```{r setup}
library(xbioclim)
```

## Input data format

Every function expects four numeric vectors of **length 12**, one value per
calendar month (January = 1, …, December = 12):

| Argument | Meaning | Units |
|----------|---------|-------|
| `tas`    | Monthly mean temperature | °C |
| `tasmax` | Monthly maximum temperature | °C |
| `tasmin` | Monthly minimum temperature | °C |
| `pr`     | Monthly precipitation | mm |

### Real-world example: Mexico City climate normals

The values below are approximate 1991–2020 climate normals for Mexico City
(19.4°N, 99.1°W, ~2240 m a.s.l.) drawn from publicly available records.

```{r mexico-city-data}
# Monthly mean temperature (°C), Jan–Dec
tas    <- c(12.0, 13.3, 15.8, 17.0, 17.5, 16.5,
            15.6, 15.6, 15.2, 14.5, 12.9, 12.0)

# Monthly maximum temperature (°C)
tasmax <- c(21.2, 23.0, 25.7, 26.6, 26.3, 24.4,
            22.8, 22.9, 22.3, 21.7, 20.5, 20.4)

# Monthly minimum temperature (°C)
tasmin <- c( 3.5,  4.5,  6.9,  9.4, 10.8, 11.4,
            10.8, 10.7,  9.9,  8.2,  5.9,  4.1)

# Monthly precipitation (mm)
pr     <- c(10,  7, 11, 21, 52, 132,
           163, 152, 116, 62, 16,   8)
```

## Computing all 19 variables at once

`bioclim()` returns a **named numeric vector** of length 19:

```{r all-19}
result <- bioclim(tas, tasmax, tasmin, pr)
print(result)
```

You can index by name or position:

```{r index}
result[["bio01"]]   # Mean Annual Temperature
result["bio12"]     # Annual Precipitation
result[c("bio04", "bio15")]  # Both seasonality indices
```

## Computing individual variables

Each variable has its own function.  Use these when you only need one or a
few variables to avoid computing the full set.

### Temperature variables

```{r temp-vars}
bio01(tas)              # BIO01: Mean Annual Temperature
bio02(tasmax, tasmin)   # BIO02: Mean Diurnal Range
bio03(tasmax, tasmin)   # BIO03: Isothermality
bio04(tas)              # BIO04: Temperature Seasonality
bio05(tasmax)           # BIO05: Max Temperature of Warmest Month
bio06(tasmin)           # BIO06: Min Temperature of Coldest Month
bio07(tasmax, tasmin)   # BIO07: Temperature Annual Range
bio08(tas, pr)          # BIO08: Mean Temp of Wettest Quarter
bio09(tas, pr)          # BIO09: Mean Temp of Driest Quarter
bio10(tas)              # BIO10: Mean Temp of Warmest Quarter
bio11(tas)              # BIO11: Mean Temp of Coldest Quarter
```

### Precipitation variables

```{r precip-vars}
bio12(pr)               # BIO12: Annual Precipitation
bio13(pr)               # BIO13: Precipitation of Wettest Month
bio14(pr)               # BIO14: Precipitation of Driest Month
bio15(pr)               # BIO15: Precipitation Seasonality (CV)
bio16(pr)               # BIO16: Precipitation of Wettest Quarter
bio17(pr)               # BIO17: Precipitation of Driest Quarter
bio18(tas, pr)          # BIO18: Precipitation of Warmest Quarter
bio19(tas, pr)          # BIO19: Precipitation of Coldest Quarter
```

## Multi-pixel (matrix) processing

When you have climate data for many locations, the simplest approach is to
store each monthly variable as a **matrix** with rows = pixels and
columns = months, then apply `bioclim()` (or individual functions) row-wise
with `apply()`.

```{r multi-pixel}
# Simulate data for 5 locations
set.seed(42)
n_pixels <- 5

make_climate <- function(n) {
  tas_base <- runif(n, 5, 20)
  matrix(
    outer(tas_base, sin(seq(0, pi, length.out = 12)) * 10, "+"),
    nrow = n, ncol = 12
  )
}

TAS    <- make_climate(n_pixels)
TASMAX <- TAS + matrix(runif(n_pixels * 12, 3, 8), nrow = n_pixels)
TASMIN <- TAS - matrix(runif(n_pixels * 12, 3, 8), nrow = n_pixels)
PR     <- abs(matrix(rnorm(n_pixels * 12, mean = 50, sd = 30), nrow = n_pixels))

# Debugging: print dimensions
print(dim(TAS))    # Debug step: expect (5, 12)
print(dim(TASMAX))
print(dim(TASMIN))
print(dim(PR))

# Apply processing
results <- t(
  vapply(seq_len(nrow(TAS)), function(i) {
    bioclim(TAS[i, ], TASMAX[i, ], TASMIN[i, ], PR[i, ])
  }, FUN.VALUE = numeric(19))
)

head(results)
```

## Complete variable reference

| Variable | Description |
|----------|-------------|
| BIO01 | Mean Annual Temperature |
| BIO02 | Mean Diurnal Range (mean of monthly tasmax − tasmin) |
| BIO03 | Isothermality (100 × BIO02 / BIO07) |
| BIO04 | Temperature Seasonality (100 × population SD of monthly tas) |
| BIO05 | Max Temperature of Warmest Month |
| BIO06 | Min Temperature of Coldest Month |
| BIO07 | Temperature Annual Range (BIO05 − BIO06) |
| BIO08 | Mean Temperature of Wettest Quarter |
| BIO09 | Mean Temperature of Driest Quarter |
| BIO10 | Mean Temperature of Warmest Quarter |
| BIO11 | Mean Temperature of Coldest Quarter |
| BIO12 | Annual Precipitation |
| BIO13 | Precipitation of Wettest Month |
| BIO14 | Precipitation of Driest Month |
| BIO15 | Precipitation Seasonality (Coefficient of Variation) |
| BIO16 | Precipitation of Wettest Quarter |
| BIO17 | Precipitation of Driest Quarter |
| BIO18 | Precipitation of Warmest Quarter |
| BIO19 | Precipitation of Coldest Quarter |

## Input validation

All functions validate their inputs and produce informative error messages:

```{r validation, error = TRUE}
bio01(1:6)                    # Wrong length
bio01(letters[1:12])          # Non-numeric
bioclim(1:12, 2:13, 0:11, 1:11)  # pr has length 11
```
