---
title: "xbioclim vs terra: Side-by-Side Comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{xbioclim vs terra: Side-by-Side Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The [terra](https://rspatial.github.io/terra/) package includes
`terra::bioclim()` which computes bioclimatic variables from a
`SpatRaster` stack.  **xbioclim** fills a complementary niche: it works
directly with plain R numeric vectors (and, by extension, matrices and
data frames), without requiring any spatial infrastructure.

This vignette walks through the key similarities and differences between
the two packages so you can choose the right tool—or combine both—for your
workflow.

```{r setup}
library(xbioclim)
```

## Shared vocabulary

Both packages follow the same **WorldClim** variable definitions (BIO01–BIO19)
and the same underlying formulas.  If you already know how to interpret the
variables from one package, the numbers produced by the other should be
directly comparable.

## Input format

| Feature | xbioclim | terra |
|---------|-----------|-------|
| Input type | Plain numeric vector (length 12) | `SpatRaster` with 12 monthly layers |
| Spatial awareness | None – pure numeric | Full CRS + extent handling |
| Multi-pixel | Row-wise `apply()` over a matrix | Operates natively on rasters |
| Installation | Lightweight (no system libs needed) | Requires GDAL/PROJ |

### xbioclim: vector inputs

```{r xbioclim-example}
# Monthly climate normals for a single location
tas    <- c(12.0, 13.3, 15.8, 17.0, 17.5, 16.5,
            15.6, 15.6, 15.2, 14.5, 12.9, 12.0)
tasmax <- c(21.2, 23.0, 25.7, 26.6, 26.3, 24.4,
            22.8, 22.9, 22.3, 21.7, 20.5, 20.4)
tasmin <- c( 3.5,  4.5,  6.9,  9.4, 10.8, 11.4,
            10.8, 10.7,  9.9,  8.2,  5.9,  4.1)
pr     <- c(10,  7, 11, 21, 52, 132,
           163, 152, 116, 62, 16,   8)

# All 19 variables in one call
result_rx <- bioclim(tas, tasmax, tasmin, pr)
result_rx
```

### terra: raster inputs (conceptual)

The equivalent terra workflow requires a `SpatRaster` where each layer holds
one month of data.  Below is a minimal reproducible sketch—no real raster
files are read from disk:

```r
library(terra)

# Assume `r_tas`, `r_tasmax`, `r_tasmin`, `r_pr` are each SpatRasters
# with 12 layers (one per month), all sharing the same CRS and extent.

# terra::bioclim() accepts the four stacks and returns a 19-layer SpatRaster
bio_raster <- terra::bioclim(r_tas, r_tasmax, r_tasmin, r_pr)

# Extract values at a single cell (equivalent to xbioclim on one pixel)
cell_id <- cellFromXY(bio_raster, cbind(-99.1, 19.4))
values(bio_raster)[cell_id, ]
```

## Function-level comparison

### Computing a single variable

**xbioclim** exposes every variable as its own named function, making
intent explicit and reducing overhead when only one variable is needed:

```{r single-var-rx}
bio04(tas)   # Temperature Seasonality – xbioclim
bio15(pr)    # Precipitation Seasonality – xbioclim
```

In **terra** you would compute all 19 layers and then subset the desired
layer from the output `SpatRaster`:

```r
# terra approach – always computes all 19
bio_raster <- terra::bioclim(r_tas, r_tasmax, r_tasmin, r_pr)
bio_raster[["bio04"]]  # Temperature Seasonality layer
```

### Computing all 19 variables

```{r all-19-rx}
result <- bioclim(tas, tasmax, tasmin, pr)
print(result)
```

In terra the output is a `SpatRaster` with the same variable names as
layer names, so downstream raster algebra works naturally:

```r
bio_raster <- terra::bioclim(r_tas, r_tasmax, r_tasmin, r_pr)
names(bio_raster)  # "bio01" ... "bio19"
```

## Applying xbioclim to tabular climate data

A common use-case outside of raster workflows is working with **station data**
or **point samples** stored in a data frame.  xbioclim is ideal here:

```{r tabular}
# Simulated station records: 3 stations × 12 months
set.seed(1)
n_stations <- 3
months <- 1:12

station_df <- data.frame(
  station = rep(paste0("S", 1:n_stations), each = 12),
  month   = rep(months, times = n_stations),
  tas     = c(
    c(5, 6, 9, 13, 17, 21, 23, 22, 18, 13, 8, 5),     # temperate
    c(24, 25, 26, 27, 28, 27, 26, 26, 26, 25, 24, 23), # tropical
    c(-10, -8, -2, 5, 12, 17, 20, 18, 12, 4, -3, -8)   # continental
  ),
  tasmax  = c(
    c(8, 10, 14, 18, 23, 27, 30, 29, 25, 19, 13, 9),
    c(30, 31, 32, 33, 34, 33, 32, 32, 32, 31, 30, 29),
    c(-5, -3,  4, 11, 18, 24, 27, 25, 18,  9,  2, -3)
  ),
  tasmin  = c(
    c(1, 2, 4, 8, 12, 16, 18, 17, 13, 8, 4, 1),
    c(20, 21, 22, 23, 24, 23, 22, 22, 22, 21, 20, 19),
    c(-16, -14, -9, -1, 6, 11, 14, 12, 6, -1, -8, -14)
  ),
  pr      = c(
    c(60, 50, 45, 40, 35, 20, 10, 15, 35, 55, 65, 65),
    c(5, 10, 20, 40, 80, 150, 180, 160, 120, 60, 20, 8),
    c(25, 20, 22, 30, 45, 60, 70, 65, 45, 30, 28, 25)
  )
)

# Apply bioclim() per station using split + lapply
# Convert station to character to ensure robust splitting
bio_list <- lapply(
  split(station_df, as.character(station_df$station)),
  function(d) bioclim(d$tas, d$tasmax, d$tasmin, d$pr)
)

# Combine into a data frame, keeping numeric columns numeric
bio_df <- do.call(rbind, lapply(names(bio_list), function(nm) {
  as.data.frame(as.list(c(station = nm, bio_list[[nm]])))
}))

print(bio_df[, c("station", "bio01", "bio04", "bio12", "bio15")])
```

## When to use each package

| Scenario | Recommended package |
|----------|---------------------|
| Processing global rasters (NetCDF, GeoTIFF) | **terra** |
| Computing variables for point/station data | **xbioclim** |
| Lightweight scripts with no spatial dependency | **xbioclim** |
| Integration into existing terra workflows | **terra** (or both) |
| Custom block processing with non-standard data | **xbioclim** |
| SDM workflows that use `terra::extract()` output | **xbioclim** |

## Bridging both packages

If you extract raster values with terra and want to compute bioclim variables
with xbioclim (e.g., for custom weighting or filtering), the workflow is:

```r
library(terra)
library(xbioclim)

# 1. Extract monthly values at species occurrence points
pts <- vect(occurrences, geom = c("lon", "lat"), crs = "EPSG:4326")
tas_vals    <- extract(r_tas,    pts)[, -1]  # drop ID column
tasmax_vals <- extract(r_tasmax, pts)[, -1]
tasmin_vals <- extract(r_tasmin, pts)[, -1]
pr_vals     <- extract(r_pr,     pts)[, -1]

# 2. Compute bioclim variables for each occurrence point
bio_matrix <- t(mapply(function(i) {
  bioclim(
    as.numeric(tas_vals[i, ]),
    as.numeric(tasmax_vals[i, ]),
    as.numeric(tasmin_vals[i, ]),
    as.numeric(pr_vals[i, ])
  )
}, seq_len(nrow(pts))))
```

This hybrid approach is particularly useful for occurrence-based niche
modelling, where you need bioclimatic values at irregular point locations
rather than over a complete raster grid.
